Density-dependent selection and the limits of relative fitness

### 2018-03-23 manuscript.tex
l

There are a variety of different measures of fitness. Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is widely used in evolutionary genetics, where the focus is on relative genotypic frequencies \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, but it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp. 178]{charlesworth_1994}). While such a connection is clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the widespread use of relative fitness in evolutionary genetics, it is important to understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep{crow_1970}. In the particular case that two types are present, the selection equation takes the familiar form
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$. Note that, in the discrete case, we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$. These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp. 468]{barton_2007}. Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would be relative log fitnesses).

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic'' growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic population growth rates. However, growth cannot continue at a constant rate indefinitely. As population density increases, crowding will cause the Malthusian parameters $r_i$ to decline (e.g. \citealt[pp. 203]{begon_1990}). It is then possible that selection is density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}. We would then need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}. In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in the presence of density-dependent selection, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.

@@

In light of the complications of density-dependence, the assignment of a density-independent relative fitness to each type can be justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

However, the preceding arguments do not imply that the constant relative fitness idealization of population genetics \textit{only} applies when selection is weak and $N$ is stable (or when selection is actually density-independent). This idealization is powerful, and so it is important to understand the specifics of how it succeeds or fails when selection is not weak, or $N$ is not stable. For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed model of selection based on Malthusian parameters or birth/death rates in this important model system? And if we make this switch, how much ecological detail do we need?

>>>>

There are a variety of different measures of fitness, such as expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is widely used in evolutionary genetics, where the focus is on relative genotypic frequencies \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, but it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp. 178]{charlesworth_1994}). While such a connection is clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the widespread use of relative fitness in evolutionary genetics, it is important to understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep{crow_1970}. In the particular case that two types are present, the selection equation takes the familiar form
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$. Note that we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$. These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp. 468]{barton_2007}. Similarly in the continuous case, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would be relative log fitnesses).

In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic growth rates. However, growth cannot continue at a constant rate indefinitely. As population density increases, the Malthusian parameters $r_i$ will eventually start to decline due to crowding (e.g. \citealt[pp. 203]{begon_1990}). When population growth is density-regulated, selection can be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}. Eq.~\eqref{eq:canonical} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonical} depends. In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in the presence of density-dependent selection, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.

@@

In light of the complications of density-dependence, the assignment of a density-independent relative fitness to each type has been justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

However, the preceding arguments do not imply that the constant relative fitness idealization of population genetics \textit{only} applies when selection is weak and $N$ is stable (or when selection is actually density-independent). The idealization of assigning relative fitness values to genotypes is powerful, and so it is important to understand the specifics of when and how it succeeds or fails when selection is not weak, or $N$ is not stable. For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed model of selection based on Malthusian parameters or birth/death rates in this important model system? And if we make this switch, how much ecological detail do we need?
